Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
Finish hacking up 4.0.  The section still uses the "seperated form" indirections, but I think it now flows nicely since the similarity between 1M splitting and 2 splitting in now more clear.

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}), splitting is also defined on 
\bbonem\ terms.
However, defining correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals), \bbonem\ terms reduce to partial values and
residual tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, the \bbonem\ term:
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
reduces (via $\redonesym$) to the residual table and partial value:
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired monostaged
outputs of splitting,
% Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
we define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mathcal{V}$,
which converts residual tables and partial values into a syntactically separated {\em masked value} of the form $\mval i q$.
%version $\mathcal{V}$.
%As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value})
The first component, $i$ of the masked value holds the result of all
first stage computations in $v$.  The second component, $q$ is a
residual representing all second stage computations in $\rtab \xi v$.

@@

Given this definition of masking, the requirements for splitting
\bbonem\ terms are as follows. For any $\colmix e A$ which reduces to
$\rtab \xi v$ (via $\redonesym$), $\rtab \xi v$ masks to $\mval i q$
(via $\vsplito$), $e$ splits into a syntactically seperated program
$\mathcal{P}$ of the form $\pipeM c l r$ (via the new relation $e
\splitonesym \mathcal{P}$) which is reduced to a masked value $\mval i
{q'}$ (via the new relation $\sepredtwosym$) where $q'$ is the same as
$q$ up to an equivalence.

%Given the definition of masking, \bbonem\ terms split via the new
%relation $e \splitonesym \mathcal{P}$, where $\mathcal{P}$ must have
%the form $\pipeM c l r$
where $c$ (called the {\em combined term}) is
a monostage term encoding all the first-stage computations in $e$, and
$l.r$ (again called the {\em resumer}) is a monostage term encoding
all the second-stage computations in $e$.

Similar to above, we also define a $\sepredtwosym$ relation 
which can be used to evaluate these seperated world \bbonep\ terms.

Given this definition of masking, we define $\sepredonesym$ by a single rule that creates masked values:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
Here we see the primary difference between world \bbtwo\ terms and world \bbonem\ terms.
Since $\coltwo e A$ reduces to an entirely second-stage result, 
its first-stage subcomputations only exist to internally pass their value to the
second stage;
but since $\colmix e A$ reduces to a multistage result, 
its first-stage subcomputations exist {\em both} to pass their value to the
second stage {\em and} to be present in the result.
Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$ reduces only to a boundary value which is passed to the resumer;
but in the $\sepredonesym$ rule, the combined term $c$ reduces to a tuple containing both the immediate result and the boundary value,
with only the latter being passed to the resumer.

A summary of these operations and the relationship between them is provided on the left side of \ref{fig:splittingSummary}.

This section proceeds by defining masking ($\vsplito$), 
then using masking to motivate the definition of \bbonem-term splitting ($\splitonesym$),
and finally coming back to define \bbtwo-term splitting ($\splittwosym$).

>>>>

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
defining correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to the monostage result of
$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, the \bbonem\ term:
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
reduces (via $\redonesym$) to the residual table and partial value:
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired monostaged
outputs of splitting,
% Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
we choose to define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mval i q$,
which converts residual tables and partial values into a syntactically separated {\em masked value}.
%version $\mathcal{V}$.
%As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value})
The first component of the masked value, $i$, is a value holding the
result of all first stage computations in $v$.  The second component,
$q$, is a residual representing all second stage computations in $\rtab
\xi v$.

@@

We define the requirements for splitting \bbonem\ terms using the
results of masking. Specifically, for any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via $\vsplito$).  Then $e$ splits into a syntactically
seperated program $\mathcal{P}$ (via the new relation $e \splitonesym
\mathcal{P}$) which is reduced to a masked value $\mval i {q'}$ (via
the new relation $\sepredtwosym$) where $q'$ is the same as $q$ up to
an equivalence.

%Given the definition of masking, \bbonem\ terms split via the new
%relation $e \splitonesym \mathcal{P}$, where $\mathcal{P}$ must have
%the form $\pipeM c l r$
Similar to \bbtwo\ term splitting, $\mathcal{P}$ takes the form
$\pipeM c l r$ where $c$ (called the {\em combined term}) is a
monostage term encoding all the first-stage computations in $e$, and
$l.r$ (again called the {\em resumer}) is a monostage term encoding
all the second-stage computations in $e$.  $\sepredonesym$ is defined
by a single rule that yields masked values:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$
generated via splitting reduces only to a boundary value which is
passed to the resumer; but in the $\sepredonesym$ rule, the combined
term $c$ reduces to a tuple containing both the first stage result $i$
and the boundary value $b$ provided as input to the resumer.  A
summary of the relationship between the results of directly evaluating
\lang\ and the results of splitting is provided at left in
\ref{fig:splittingSummary}.

In the following subsections we formally define the masking
transformation ($\vsplito$), then describe the algorithm for splitting
both \bbonem\ terms ($\splitonesym$) and \bbtwo\ terms
($\splittwosym$).
